\chapter{Imperative programming}

Consider doing the laundry. If we have a common, house-hold washing machine at
our disposal --- the \emph{procedure}, may look something like this:

\begin{enumerate}

\item split the laundry into whites and colours;

\item wash whites;

\item wash colours;

\item dry the laundry;

\item neatly stack up the laundry.

\end{enumerate}

In a conventional functional programming language, such as SML, this may end up
looking like this:

\begin{code}
(stack (dry (wash (split laundry))))
\end{code}

While comprehensible, this does not read very well. Most languages we're used
to, i.e. \wikipedia{Indo-european}{Indo-European languages}, read
left-to-right, top-to-bottom. Here, the flow of the program is expressed
right-to-left, and there is no notion of top-down descent at all. If anything,
there's a notion of bottom-up ascent, if we for instance, lay out the program
like this:

\begin{code}
(stack
  (dry
    (wash
      (split laundry))))
\end{code}

This reads like we're putting the cart before the horse. Indeed, some tasks ---
such as, doing the laundry, cooking a meal, writing a program, etc. --- are
inherently sequential. Functional programming languages can be
\explain{syntactically}{A programming language has syntactic rules defining the
valid positioning of symbols. In this case, the syntactic rule is that the name
precedes the arguments in a function call.} unfit for writing such programs
well.

We will develop the notion of a \emph{well-written program} throughout these
lecture notes, but we begin with the following definition:

\begin{definition}

A well-written program is a well-read program.

\end{definition}

That is, a well-written program is comprehensible to another programmer.
Naturally, not any old programmer will do, but someone with relative knowledge
of the problem domain, and programming experience comparable, or superceeding
yours, should be able to comprehend and evaluate your programs. The intent is
to facilitate a the review and evolution of your programs by your peers.

We were also somewhat \emph{declarative} when we wrote the functional program.
For instance, we've left out the fact that we should wash whites before we wash
colours. Actually, we didn't even mention that we split the laundry into whites
and colours; we've left it up to the \var{wash} and \var{split} function
definitions to discuss such ``details''.

\begin{definition}

A declarative style of programming is a style where we tell the reader ``what''
we're doing, rather than ``how'' we're doing it.

\end{definition}

An alternative functional program could be:

\begin{code}
let
  (dirtyWhites, dirtyColours) = split dirtyLaundry
  cleanWetWhites = wash dirtyWhites
  cleanWetColours = wash dirtyColours
  cleanLaundry = dry (cleanWetWhites, cleanWetColours)
in
  stack cleanLaundry
\end{code}

We'll refer to lines 1--7 as a ``\mono{let}-block'', lines 2--4 as a
``\mono{let}-definitions'', and line 7 as a ``\mono{let}-expression''.

In functional programming, the \mono{let}-syntax can be used to reduce the
complexity of an expression by giving intuitive names to subexpressions. This
allows the reader to read the expression, and delve into the details of its
constituents only if the reader wishes to do so. As long as important deails
are retained, this is a useful technique for making your programs more
readable.

In this particular case, the expression reduces to simply \code{stack
cleanLaundry}. This seems overly declarative. We're completely disregarding the
important aspects of ``doing the laundry''. We're so eager not to tell the
reader ``how'' we're doing it, that we forget to tell the reader ``what'' we're
doing.

The \mono{let}-syntax does however provide for a nice segway into another
\explain{programming paradigm}{A way of writing a particular class of programms
well.}, more suitable for writing out \emph{procedures}. In particular, the
\mono{let}-syntax provides more than a simple mechanism for naming
subexpressions. Notice how we could use the name \var{dirtyWhites} when we
defined \var{cleanWetWhites}, or how we could use \var{cleanWetWhites} when we
defined \var{cleanLaundry}.

It is as if some sort of ``state'' is retained from one defintion to the next,
such that names defined further up can be used in definitions further down. If
we represent ``state'' as the set of names available for use in an expression
we can specify how the state develops throughout the \mono{let}-block: 

\begin{code}
let
  {dirtyLaundry}
  {dirtyLaundry, dirtyWhites, dirtyColours}
  {dirtyLaundry, dirtyWhites, dirtyColours,
    cleanWetWhites}
  {dirtyLaundry, dirtyWhites, dirtyColours,
    cleanWetWhites, cleanWetColours}
in
  {dirtyLaundry, dirtyWhites, dirtyColours,
    cleanWetWhites, cleanWetColours, cleanLaundry}
\end{code}

It seems overly excessive to have the state at the \mono{let}-expression (lines
9--10) contain all these ``intermediate'' names, such as \var{dirtyWhites},
\var{cleanWetColours}, etc. All we really want in the \mono{let}-expression is
the \var{cleanLaundry}. What's more, it does not conceptually make sense to be
able to use the dirty, or wet laundry once the laundry has been washed and
dried, right? Instead, we'd like for the state develop in a fashion similar to
the following:

\begin{code}
{dirtyLaundry}
{dirtyWhites, dirtyColours}
{cleanWetWhites, cleanWetColours}
{cleanDryLaundry}
\end{code}

While we could attempt to fix this using nested \mono{let}-blocks or
\explain{auxiliary functions}{Functions whos sole purpose is to make other
functions easier to comprehend.}, the final code will not be as easy as
1-2-3-4-5, i.e. as simple as our initial specification of the procedure.

The thing that is getting in the way here is the functional paradigm we're
dealing with \emph{immutable} values. That is, values that can be created and
destroyed, but not modified. Indeed, if the pile of laundry was a
\emph{mutable} value, we could simply step-wise modify that value from a dirty
pile to a clean stack. After-all, this resembles what we do with the laundry in
real life.

An alternative programming paradigm, the \emph{imperative}, allows us to write
the program like this:

\begin{code}
splitDirtyLaundry(laundry)
washWhites(laundry)
washColours(laundry)
cleanWetColours = wash(dirtyColours)
dry(laundry)
stack(laundry)
\end{code}

Here, \var{laundry} is a value that begins as a pile of dirty laundry. In step
1 it is \emph{transformed} into another value: a pair of piles --- whites and
colours. At this point, we can think of \var{laundry} as a container, say a
two-compartment basket. We can then access the piles in the respective
compartments using what is known as the \emph{dot-syntax}, i.e.
\var{laundry.whites} and \var{laundry.colours}. After the laundry is washed, it
is put back in the same compartment. In step 4, we pick up the laundry from
both compartments and hang it up to dry. The laundry is thus transformed into a
clean, dry pile of laundry. In the last step, we simply take this pile and
neatly stack it. The resulting \var{laundry} is hence a neat stack of clean,
dry laundry.

\begin{definition}
An immutable value does not change.
\end{definition}

In the functional paradigm, we're programming by defining mathematical
functions. Functions themselves are defined in terms of applications of other
functions, and branching based on the outcomes of boolean predicates.

An alternative programming paradigm, the \emph{imperative}, 

{\bf Assignment}

ASCII report.

The DIKU canteen is the only completely student-driven canteen on the
Copenhagen University campus. The canteen was established in 1971, and runs
till this day. As a student-driven canteen, even the accounting is done by
students. No biggie for Datalogy students, eh? We'll see.

It is possible to buy from the canteen in bulk. The usual order consists of
perhaps 30 bottles of soft drinks, and 30 bottles of beer. Although all soft
drinks have the same price, beer prices may vary depending on how enthusiastic
you are about beer. Indeed, DIKU alumni and research groups can be quite a
picky bunch when it comes to beer specialities. We can without loss of
generality constrain ourselves to beer.

Every time a bulk purchase is made, the items and quantities purchased are
noted in a journal, leaving it to accounting to handle later.

This is where you come in. We need to write a program that can process the
journal and produce a summary report in ASCII. The summary report should
process the entire journal and provide as output:

\begin{itemize}

\item the total number of times each type of beer was purchased;

\item the total revenue made from each type of beer;

\item the overall revenue from bulk purchases;

\item an overview of the development of the revenue over time.

\end{itemize}

\begin{verbatim}
Guld Tuborg 10
Grøn Tuborg 8
Leffe Brun 15
Leffe Blond 15
\end{verbatim}

The users will
sometimes however chose to buy abnormal quantities of various drinks.

- Canteen accounting. List of items with their prices, followed by event
  receipts.

- ASCII bar chart.

- Challenge of the week: ASCII line chart. Hint: don't drink and derive!

Tracing the attacker's IP.

- GUI in visual basic, LOL.

- Parse (csv-based) log files.

- Find comon behaviour:

- - Periodic access to the same resource.

- - Frequent access to different resources.
