\chapter{Imperative, Procedural and Structured Programming}

Consider doing the laundry. If we have a common, house-hold washing machine at
our disposal, the \emph{procedure} may look something like this:

\begin{code}
split the laundry into whites and colours;
wash whites;
wash colours;
dry the laundry;
neatly stack up the laundry.
\end{code}

In a conventional functional programming language, such as SML, this may end up
looking like this:

\begin{code}
(stack (dry (wash (split laundry))))
\end{code}

While comprehensible, this does not read very well. Most languages we're used
to, i.e. \wikipedia{Indo-euoropean}{Indo-European} languages, read
left-to-right, top-to-bottom. Here, the flow of the program is expressed
right-to-left, and there is no notion of top-down descent at all. If anything,
there's a notion of bottom-up ascent, if we for instance, lay out the program
like this:

\begin{codebox}
\li (stack
\zi   (dry
\zi     (wash
\zi       (split laundry))))
\end{codebox}

So we ``stack the dried and washed laundry''? This reads like we're putting the
cart before the horse, and that we're trying to say too much in one breath.
Some tasks, such as doing the laundry or cooking a meal, are inherently
sequential, i.e. they are done by performing a sequence of more basic tasks,
sometimes quite a lengthy sequence.  Functional programming languages can be
\explain{syntactically}{A programming language has syntactic rules defining the
valid positioning of symbols. In this case, the syntactic rules are that the
name precedes the arguments in a function call, and that the body of a function
is a single expression.} unfit for writing programs to perform such tasks
\emph{well}.

% TODO: the segway is gone.

We will develop the notion of a \emph{well-written program} throughout these
lecture notes, but we begin with the following definition:

\begin{definition}

A well-written program is a well-read program.

\end{definition}

That is, a well-written program is comprehensible to another programmer.
Naturally, not any old programmer will do, but someone with relative knowledge
of the problem domain, and programming experience comparable, or superceeding
yours, should be able to comprehend and evaluate your programs. The intent is
to facilitate the development of peer-reviewed programs.

% TODO: Perhaps continue frome here?

We were also somewhat \emph{declarative} when we wrote the functional program.
For instance, we've left out the fact that we should wash whites before we wash
colours. Actually, we didn't even mention that we split the laundry into whites
and colours; we've left it up to the \var{wash} and \var{split} functions to
discuss such ``details''.

\begin{definition}

A declarative style of programming is a style where we tell the reader ``what''
we're doing, rather than ``how'' we're doing it.

\end{definition}

An alternative functional program could be:

\begin{code}
let
  (dirtyWhites, dirtyColours) = split dirtyLaundry
  cleanWetWhites = wash dirtyWhites
  cleanWetColours = wash dirtyColours
  cleanLaundry = dry (append cleanWetWhites cleanWetColours)
in
  stack cleanLaundry
\end{code}

We'll refer to lines 1--7 as a ``\mono{let}-block'', lines 2--4 as a
``\mono{let}-definitions'', and line 7 as a ``\mono{let}-expression''.

In functional programming, the \mono{let}-syntax can be used to reduce the
complexity of an expression by giving intuitive names to some of its
constituents. This lets the reader read a simpler expression, and delve into
the details only if necessary. So long as the important details are retained in
the \mono{let}-expression, this is a useful technique for making your programs
more readable.

However, in this particular case, the expression reduces to the rather mundane
``\code{stack cleanLaundry}''. This seems overly declarative. We're completely
disregarding the important aspects of ``doing the laundry''. We're so eager not
to tell the reader ``how'' we're doing it, that we forget to tell the reader
``what'' we're doing.

The \mono{let}-syntax does however provide for a segway to another programming
paradigm, more suitable for writing procedures like this.

\begin{definition}

A programming paradigm is a particular way of writing a certain class of
programs well.

\end{definition}

In particular, the \mono{let}-syntax provides more than a simple mechanism for
naming subexpressions. Notice, how we could use the name \var{dirtyWhites} when
we defined \var{cleanWetWhites}, or how we could use \var{cleanWetWhites} when
we defined \var{cleanLaundry}.

It is as if some sort of ``state'' evolves from one defintion to the next, in
the sense that a name defined further up can be used in definitions further
down. If we represent this ``state'' as the set of names available in the
context of an expression, we can specify how the state develops throughout the
\mono{let}-block: 

\begin{codebox}
\li let
\li   {dirtyLaundry}
\li   {dirtyLaundry, dirtyWhites, dirtyColours}
\li   {dirtyLaundry, dirtyWhites, dirtyColours,
\zi     cleanWetWhites}
\li   {dirtyLaundry, dirtyWhites, dirtyColours,
\zi     cleanWetWhites, cleanWetColours}
\li in
\li   {dirtyLaundry, dirtyWhites, dirtyColours,
\zi     cleanWetWhites, cleanWetColours, cleanLaundry}
\end{codebox}

It seems overly excessive to have the state at the \mono{let}-expression (line
7) contain various intermediate names, such as \var{dirtyWhites},
\var{cleanWetColours}, etc. All we really want in the \mono{let}-expression is
the \var{cleanLaundry}. What's more, it does not conceptually make sense to be
able to use the dirty, or wet laundry once the laundry has been washed and
dried, right? Indeed, we'd like for the state develop in a fashion similar to
this:

\begin{code}
let
  {dirtyLaundry}
  {dirtyWhites, dirtyColours}
  {cleanWetWhites, cleanWetColours}
in
  {cleanDryLaundry}
\end{code}

While we could attempt to attain this using nested \mono{let}-blocks or
\explain{auxiliary functions}{Functions whos sole purpose is to make other
functions more readable.}, the code will not be as easy as 1-2-3, i.e.  as
simple as our initial specification of the procedure.

The thing that is getting in the way here is that in the functional paradigm
we're dealing with \emph{immutable} values. That is, values that can be created
and destroyed, but not modified. Indeed, if the pile of laundry was a
\emph{mutable} value, we could simply step-wise modify that value from a dirty
pile to a clean stack. After-all, this resembles what we do with the laundry in
real life.

\begin{definition}

A mutable value is a type of value that can have multiple concrete values
throughout its lifetime.

\end{definition}

Let us explore the benfit of mutable values by considering the first
\mono{let}-definition:

\begin{code}
dirtyWhites, dirtyColours = split dirtyLaundry
\end{code}

If we let \var{dirtyLaundry} initially be a basket of mixed dirty laundry, we
can let the \emph{effect} of this statement be that \var{dirtyWhites} becomes a
pile of dirty whites, and \var{dirtyColours} becomes a pile of dirty colours,
while \var{dirtyLaundry} becomes an \emph{empty} basket. That is, we can have
the laundry \emph{moved} from the basket to the respective piles. Again, this
resembles what we do in real life.

In the context of mutable values, it no longer makes sense to represent
``state'' as the mere set of names available in the context of an expression.
It now also matters which concrete value a name refers to in the context of an
expression. Indeed, before the laundry is split, \var{dirtyLaundry} is a basket
full of dirty clothes; after it, it is an empty basket. If we represent baskets
and piles as good old lists, we can define the effect of the split procedure as
the change of state from

\begin{codebox}
\li { dirtyLaundry = [white, colour, white, colour, ...] }
\end{codebox}

to

\begin{codebox}
\li {
\zi   dirtyLaundry = [],
\zi   dirtyWhites = [white, white, ...],
\zi   dirtyColours = [colour, colour, ...]
\zi }
\end{codebox}

Let us now consider the split procedure itself:

\begin{codebox}
\li instanciate a whites pile;
\li instanciate a colours pile;
\zi  
\li if the basket is empty, end the procedure;
\zi  
\li pick up a piece of laundry from the basket;
\li if the piece is white, throw it in the whites pile;
\li otherwise, throw it in the colours pile;
\zi  
\li go to 3.
\end{codebox}

In this specification we've used the word ``instanciate''. In the context of
doing the laundry, this simply means finding a large enough empty spot on the
floor. Once we've found such a spot, we'd like to keep that spot in mind, so
that we know where to throw pieces of laundry after we pick them up from the
basket and inspect them. We hence retain a \emph{reference} to this spot.
After the steps 1--2 our state looks like this:

\begin{codebox}
\li {
\zi   dirtyLaundry = [white, colour, white, colour, ...],
\zi   dirtyWhites = [],
\zi   dirtyColours = []
\zi }
\end{codebox}

Notice, how \var{dirtyLaundry} is also merely a reference.

Steps 3--7 define an \emph{iterative} procedure, in the sense that the last
thing that happens in the procedure is a jump back its beginning.

Indeed, in SML, we could define this procedure as follows:

\begin{code}
split [] dirtyWhites dirtyColours =
  (dirtyWhites, dirtyColours)
split (piece :: dirtyLaundry) dirtyWhites dirtyColours =
  if isWhite piece
  then
    split dirtyLaundry (piece :: dirtyWhites) dirtyColours
  else
    split dirtyLaundry dirtyWhites (piece :: dirtyColours)
\end{code}

Although most imperative programming languages have retained the \var{goto}
statement, it is considered harmful. Instead, programmers are encouraged to use
various built-in \emph{control structures}, i.e. built-in higher-order
functions with native language support.

Wrt. defining iterative procedures, imperative languages often come with a
series of \emph{loop} control structures.

\begin{definition}

A loop control structure repeats a sequence of steps.

\end{definition}

One such control structure is the \emph{while} loop.

\begin{definition}

A while loop control structure repeats a sequence of steps as long as a boolean
condition is met.

\end{definition}

Lines 3--7 of the split procedure can be rewritten using a while loop in the
following fashion:

\begin{code}
while (basket is not empty)
{
  pick up a piece of laundry from the basket;
  if the piece is white, throw it in the whites pile;
  otherwise, throw it in the colours pile;
}
\end{code}
